---
title: "R Notebook"
output: html_notebook
---

## Exploratory Data Analysis

```{r}

library(tidyverse)
library(dplyr)
library(ggplot2)

# Visualizing distributions for categorical variables
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))

# Calculating frequency of the categorical variable
diamonds %>%
  count(cut)

# Visualizing distributions for continuous variables
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.6, color = "blue")

# Histogram for diamonds with carat size < 2
smaller <- diamonds %>% 
  filter(carat < 2)

ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.1)

# Overlaying multiple histograms for 'cut' using geom_freqpoly
ggplot(data = smaller, mapping = aes(x = carat, colour = cut)) +
  geom_freqpoly(binwidth = 0.1)
```

## Outliers/Unusual values

```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)

# Evidence of the outliers is the large extended x-axis having values that are almost not visible. We can zoom into this using coord_cartesian()
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 40))

# The values between 30-40 and 40-50 are visible now that the y-axis has limits on it. We can pick out the exact values by filtering
unusual <- diamonds %>% 
  filter(y < 3 | y > 20) %>% 
  select(price, x, y, z) %>%
  arrange(y)

unusual
```

The unusual values are 31.8 and 58.9

## Missing Values

```{r}
# Replacing unusual values with missing values
diamonds2 <- diamonds %>% 
  mutate(y = ifelse(y < 3 | y > 20, NA, y))

ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_point(na.rm = TRUE)

#Comparing scheduled departure time for cancelled and non-cancelled flights

nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(sched_dep_time)) + 
    geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4)
```

## Covariation

```{r}

# Plotting frequency polygon
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)

```

From the plot, it appears that the lowest quality diamonds have the highest average price. We will create a boxplot to further analyze this

```{r}

ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_boxplot()
```

The results support the earlier finding that better quality diamonds are cheaper on an average

```{r}

#Examining mpg data
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot()

#Reordering class on the basis of median value of hwy
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy))

# Making the chart more readable by flipping the x-and y-axes
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) +
  coord_flip()
```

## Covariation between 2 categorcal variables

```{r}

ggplot(data = diamonds) +
  geom_count(mapping = aes(x = cut, y = color))

# The size of the circle in the plot depicts number of observations occurring at at each combination of values  

# Computing covariation using dplyr and geom_tile
diamonds %>% 
  count(color, cut)

diamonds %>% 
  count(color, cut) %>%  
  ggplot(mapping = aes(x = color, y = cut)) +
    geom_tile(mapping = aes(fill = n))

```

## Covariation between 2 continuous variables

```{r}

ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))

# For better readability in case of overplotting, we adjust transarency
ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price), alpha = 1 / 100)

# We can also use bins to show the data. Binning one continuous variable to act like a categorical variable
ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)))

```

## Patterns and Models

```{r}

# Scatterplot of Old Faithful eruptions to show patterns
ggplot(data = faithful) + 
  geom_point(mapping = aes(x = eruptions, y = waiting))

library(modelr)

# This dampens the strong relationship between carat and price
mod <- lm(log(price) ~ log(carat), data = diamonds)

diamonds2 <- diamonds %>% 
  add_residuals(mod) %>% 
  mutate(resid = exp(resid))

ggplot(data = diamonds2) + 
  geom_point(mapping = aes(x = carat, y = resid))

# We can now test the relationship between cut and price
ggplot(data = diamonds2) + 
  geom_boxplot(mapping = aes(x = cut, y = resid))

```

This tells us relative to their size, better quality diamonds are more expensive.
