---
title: "R Notebook"
output: html_notebook
---

## Exploratory Data Analysis

```{r}
# Setting working directory
setwd("C:/Users/Sanjana/Documents/GitHub/ANLY506/CodePortfolio/")

# Returning current working directory
getwd()

library(readr)
library(dplyr)

# Importing data by reading csv file
ozone <- read_csv("US EPA data 2017.csv")

# Retrieving column names of the data
names(ozone)

#Re-writing names to remove spaces
names(ozone) <- make.names(names(ozone))

names(ozone)
```


## Checking the data package

```{r}

nrow(ozone)

ncol(ozone)

#Checking the data structure
str(ozone)

#Checking the top and bottom 10 values of the data
head(ozone, n=10)

tail(ozone, n=10) # this helps identify if there is a problem with dataat the bottom of the dataset

# Checking number of unique states
select(ozone, State.Name) %>% unique %>% nrow
```

Since we know there are only 50 US states, we take a look at the data entries

```{r}

unique(ozone$State.Name)

```

We can see from the list of unique entries that the extra states here are Washington D.C., Puerto Rico, Virgin Islands and Country of Mexico

## Data Analysis

```{r}

# Retrieving data table based on filters
filter(ozone, Observation.Percent == "97") %>% 
         select(State.Name, City.Name, Local.Site.Name, Observation.Count)

#Taking a look at the data
summary(ozone$Observation.Percent)
```

 We can see that there is a wide range in the observation percentages recorded, but the median remains high at 97 percent 

We want to know which counties have the highest observation percent

```{r}
ranking <- group_by(ozone, State.Name, County.Name) %>%
         summarize(ozone = mean(Observation.Percent)) %>%
         as.data.frame %>%
         arrange(desc(ozone))

head(ranking,10)

# We look at the 10 lowest counties too
tail(ranking,10)

#Next we focus on the county with the highest observation percentage
filter(ozone, State.Name == "Arkansas" & County.Name == "Jackson") %>% nrow
```

