---
title: "Week3"
author: "Sanjana"
date: "June 6, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Week 3: Importing, savng and managing data


```{r, warning = FALSE, message=FALSE, results="hide"}
# Print my current working directory
getwd()

# Change my working directory to the following path
setwd(dir = "C:/Users/Sanjana/Documents/GitHub/ANLY506/CodePortfolio")

```

## Printing objects in workspace


```{r, warning = FALSE, message=FALSE, results="hide"}

# Print all the objects in my workspace
ls()

```

## Creating objects


```{r, warning = FALSE, message=FALSE, results="hide"}

# Create some objects that we'll save later
study1.df <- data.frame(id = 1:5, 
                        sex = c("m", "m", "f", "f", "m"), 
                        score = c(51, 20, 67, 52, 42))

score.by.sex <- aggregate(score ~ sex, 
                          FUN = mean, 
                          data = study1.df)

study1.htest <- t.test(score ~ sex, 
                       data = study1.df)

```

## Saving objects in a .RData file 

```{r, warning = FALSE, message=FALSE, results="hide"}

# Save two objects as a new .RData file in the data folder of my current working directory
save(study1.df, score.by.sex, study1.htest,
     file = "study1.RData")

```

## Saving objects in a .RData file 

```{r, warning = FALSE, message=FALSE, results="hide"}

# Save two objects as a new .RData file in the data folder of my current working directory
save(study1.df, score.by.sex, study1.htest,
     file = "study1.RData")

```


## Saving all objects in a .RData file 

```{r, warning = FALSE, message=FALSE, results="hide"}

# Save my workspace to complete_image.RData in the data folder of my working directory
save.image(file = "projectimage.RData")

```
  
## Loading all objects in a .RData file to current workspace 

```{r, warning = FALSE, message=FALSE, results="hide"}

# Load objects in study1.RData into my workspace
load(file = "study1.RData")

# Load objects in study1.RData into my workspace
load(file = "projectimage.RData")

```

## Removing objects from workspace 

```{r, warning = FALSE, message=FALSE, results="hide"}

rm(study1.htest)


```

## Saving dataframe as txt file

```{r, warning = FALSE, message=FALSE, results="hide"}

# Write the study1 dataframe object to a tab-delimited text file called Earnings.txt in my working directory

write.table(x = study1.df,
            file = "Study1.txt",  # Save the file as Study1.txt
            sep = "\t")            # Make the columns tab-delimited


```


## Saving Study1 txt file as a new object mydata

```{r, warning = FALSE, message=FALSE, results="hide"}

# Read a tab-delimited text file called Study1.txt from the data folder in my working directory into R and store as a new object called mydata

mydata <- read.table(file = 'Study1.txt',       # file is in a data folder in my working directory
                     sep = '\t',                  # file is tab--delimited
                     header = TRUE,               # the first row of the data is a header row
                     stringsAsFactors = FALSE)    # do NOT convert strings to factors!!
```



## Reading from web URL

```{r, warning = FALSE, message=FALSE, results="hide"}

fromweb <- read.table(file = 'http://goo.gl/jTNf6P',
                      sep = '\t',
                      header = TRUE)

# Print the result
fromweb

```